---
title: "reprex-perarson-distro"
author: "Maximilian Held"
date: "22 January 2016"
output: html_document
---

I want to compare *observed* bivariate (Pearson's $\rho$ and Spearman's $\rho$) correlations coefficients with what would be expected from random data.

Assume that we measure, say, 36, cases across *very* many variables (1000).
(I know this is odd, it's called [Q methodology](https://en.wikipedia.org/wiki/Q_methodology).
Assume further that *each of the variables* is (strictly) normally distributed  *across the cases*.
(Again, very odd, but true because people as *people-variables* rank order *item-cases* under a normal distribution.)

So, *if people sorted randomly*, we should get:

```{r dataset}
m <- sapply(X = 1:1000, FUN = function(x) rnorm(36))
```

Now -- because this is Q methodology -- we correlate *all people-variables*:

```{r cor}
cors <- cor(x = m, method = "pearson")
```

Then we try to plot that, and superimpose the *distribution* of Pearson's correlation coefficient in random data, which should actually be quite close to the observed correlations in our fake data:

```{r plot}
library(ggplot2)
cor.data <- cors[upper.tri(cors, diag = FALSE)]  # we're only interested in one of the off-diagonals, otherwise there'd be duplicates
cor.data <- as.data.frame(cor.data)  # that's how ggplot likes it
summary(cor.data)
str(cor.data)
colnames(cor.data) <- "pearson"
pearson.p <- function(r, n) {
  pofr <- ((1-r^2)^((n-4)/2))/beta(a = 1/2, b = (n-2)/2)
  return(pofr)
}
g <- NULL
g <- ggplot(data = cor.data, mapping = aes(x = pearson))
g <- g + xlim(-1,1)  # actual limits of pearsons r
g <- g + geom_histogram(mapping = aes(y = ..density..))
g <- g + stat_function(fun = pearson.p, colour = "red", args = list(n = nrow(m)))  # distro as per http://stats.stackexchange.com/questions/191937/what-is-the-distribution-of-sample-correlation-coefficients-between-two-uncorrel?noredirect=1#comment365877_191937
g
```

I remember (vaguely) that the t-distribution is relevant in this context, but I can't wrap my head around how to parametrize it properly.
In particular, are the degrees of freedom given by the number of correlations (1000^2/2-500), or the number of observations on which these correlations are based (36)?

Either way, the superimposed curve in the above is clearly wrong.

I'm also confused because, the probability distribution of Pearson's r would need to be bounded (there are *no* values beyond (-) 1) -- but the t-distribution is *not* bounded.

Which distribution describes Pearson's r in this case?

---

Bonus:

The above data are actually idealized: in my real Q-study, people-variables actually have very few columns under a normal distribution to sort their item-cases into, like so:

In effect, people-variables are actually *rank-ordering* item-cases, so Pearson's is not applicable.
As a rough-and-dirty fix, I've opted for Spearman's rho, instead.
Is the probability distribution the same for Spearman's rho?


