---
title: "Grid Generation and Plotting"
author: "Maximilian Held"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
bibliography: held_library.bib
vignette: >
  %\VignetteIndexEntry{Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

***Notice: all visualisations in the below are currently dysfunctional on pensieve.maxheld.de due to a [bug in another package ](https://github.com/maxheld83/pensieve/issues/388)***.

Closed sorts (or Q-sorts) are accomplished using 2-dimensional grids, usually for ranking items across the (horizontal) x-axis, with the (vertical) y-axis used to stack ties.^[In this sense, strictly speaking, the y-dimension of the grid is often meaningless.]

*pensieve* records such grids as `psGrid`s, simple 2-dimensional matrices, where each cell is either `TRUE` or `FALSE` depending on whether the cell is allowed.
It can be constructed by providing an arbitrary matrix.

```{r}
library(pensieve)
m <- matrix(data = c(FALSE, TRUE, TRUE, TRUE, FALSE, TRUE), nrow = 2)
grid <- psGrid(grid = m)
grid
```

In Q methodology, these grids are often specified in a shorthand way, by giving a *vector* of column heights (number of allowed ties) from the left to the right.
This vector of heights --- somewhat akin to a histogram --- is sometimes also known as a *Q distribution*.
*pensieve* offers a convenient method to coerce such vectors into `psGrid`s.


## Vector Notation

```{r}
grid <- as_psGrid(obj = c(1,2,4,5,4,2,1))
grid
```

The shorthand vector notation implies that there are never any *disallowed* cells at the bottom of the grid, or anywhere in the middle of other allowed cells.
This makes sense when using rectangular grids with the allowed ties approximating a normal distribution, as is often the case.

There is however no reason to assume that *all* sorts follow such strictures, and the logical `psGrid` is a more flexible generalisation. 


## Forced and Free

A `psGrid` implies two pieces of information about a sorting study:

1. It specifies the *maximum* possible ranks (`r ncol(grid)` for the above example) and number of ties (`r nrow(grid)` for the above example) of a sort.
2. In conjunction with the number of available items (such as from `psItems`), it specifies whether a distribution is *forced* or *free*.
    Under a forced distribution, there will be exactly as many allowed cells (`TRUE`, `r sum(grid)` in the above example) as there are items, *forcing* participants to occupy every cell.
    Under a free distribution, there will be more cells than items, and participants will be flexible in which of the allowed cells they actually use.

Distributions are sometimes only reported for forced distributions, but it should be noted that *all* closed sorts use a grid.
Even under a *completely* free distribution (with all cells `TRUE`), there is a *maximum* number of ranks and ties, if simply implied by the size of the table used.

Most *pensieve* functions do not require a `psGrid`, but the design choices it stores can be helpful for validating data and enhance analysis and output methods.
Specifying a `psGrid` is highly recommended.

For optional arguments, see `??psGrid`.


## Print Method

*pensieve* offers a `knit_print()` method to visualise `psGrid`s from inside knitr chunks.
You can type `knit_print(grid)`, or simply `grid` as in the above chunks.
Currently only output *from* chunks (not inline) and *to* HTML-outputs (not PDF, Word or others) is supported.

See help for details.
