---
title: "Item Generation and Sampling"
author: "Maximilian Held"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: held_library.bib
---

Good, evocative items lie at the heart of a Q study.
In fact, akin to the old computer science adage of  *"garbage-in / garbage-out"*, the quality of the *items* strictly limits the insights that may be gleaned from the entire study.

Generating and selecting good items is at least half science, half tradecraft.
There are some standards and frameworks that can structure your item generation and sampling, but -- for now at least -- it still requires a lot of experience, intuition and knowledge of the domain in question.

`pensieve` supports this stage of a Q study in a number of ways:

1. Items **generation and selection** can be fully documented inside of `pensieve`, making the process transparent, reproducible and easy to iterate over.
2. Items can be **professionally typeset**, and stored in `pensieve`, so that participants and researchers alike both see the items in the *same, definitive form*, be they text or images.
3. Arbitrary **additional information** can be stored alongside the items for later "R-way" analysis.

Items and related information are stored in a domain-specific format in `pensieve`, a tibble of S3 class `psItems`.
You can use the package entirely without `psItems`, though providing additional information about the items is highly recommended.
For example, providing the full item text (or image) enhances some of the downstream plots.


## `psItemContent`

`psItemContent` is the most important element of `psItems`.
It stores the **full wording** of items, or, in the case of image items or other assets the **paths to the binary sources**.
`psItemContent` is easily created from a character vector.

If it is named, these names are retained as **item handles**.
Such handles provide a succinct way to refer to items, and can also be displayed in plots and tables.
In contrast to the conventionally used statement numbers ("`sta_27`"), I recommend short, meaningful labels.
The handles are *researcher-facing*, they are *not* part of the items and are never shown to participants, lest they influence the sorting.

For an example, let's take the Lipset items from @Brown1980, with some hand-crafted item handles.

```{r}
library(pensieve)
lipset <- NULL
lipset$items <- qmethod::lipset$ltext$text
names(lipset$items) <- c(
  "accept_improvements_lower_classes",
  "expected_2_improve",
  "success_resented",
  "expect_fair_treatment",
  "no_revolutionary_inclinations",
  "moderation",
  "retain_high_place",
  "wealth_deserves",
  "eliminate_privileged",
  "accept_aristocratic",
  "govnm_secrets_ok",
  "no_secrets",
  "complete_4_success",
  "social_distinction",
  "vigilantism",
  "close_2_uk",
  "prefer_voluntary_help",
  "no_wealth_for_wealth",
  "promoting_underdog",
  "like_welfare_state",
  "race_4_success",
  "corruption_accepted",
  "dif_laws_rich_poor",
  "lack_of_respect_police",
  "trust_in_police",
  "self_made",
  "respect_elite",
  "middle_of_road",
  "afterlife",
  "civil_liberties",
  "virtue",
  "raise_depressed",
  "get_ahead"
  )
lipset$items <- psItemContent(items = lipset$items, lang = "en-US")
```

You'll notice that `psItemContent()` allows you to specify a **language**, which can improve the typesetting results later on.
To learn more about acceptable language codes, see the help for `psItemContent()`.

Now that we have our canonical `psItemContent()`, what can we do with it?


### `knit_print()`

First off, we can easily **print items inside knitr documents**, by using the method for `knitr::knit_print()`.
You can do this by placing an calling `knit_print()` on some item in a knitr chunk:

```{r}
library(knitr)
knit_print(lipset$items["virtue"])
```

You can also omit the `knit_print()` call; it is default behavior for knitr to call it on returned objects.

```{r}
lipset$items["moderation"]
```

This will place the item in a quotation environment, as you can see above.

Alternatively, you can also refer to items **inline**, as you would with an in-text citation.
In this case, the *item handle* will be displayed in your document, with a footnote including the full item wording.
To do this, just insert inline R into your prose, such as `` `r
lipset$items["promoting_underdog"]` ``.

In the output, you will see item `r lipset$items["promoting_underdog"]` in the place of the above inline code, complete with a footnote automatically included.
If the output format supports this, the footnote will also be hyperlinked.

You can also `knit_print()` several items at once.

```{r}
lipset$items[c("civil_liberties", "self_made")]
```

This also works inline, for example `` `r
lipset$items[4:5]` `` yields items `r lipset$items[4:5]`.

There are some limitations:

- Line breaks, new paragraphs and other advanced markdown formatting is ignored inside the footnotes when `knit_print()` in inline.^[This would break the underlying pandoc markdown syntax for [pandoc `inline_notes`](https://pandoc.org/MANUAL.html#footnotes).]
- If you refer to the *same* item multiple times inline, there will be multiple (redundant) footnotes.
  Our `knit_print()` method does not keep track of multiple mentions of items.
  Calling item `r lipset$items["promoting_underdog"]` twice `r lipset$items["promoting_underdog"]` causes two footnotes with the same content.
- Unnamed items are simply reproduced without title, and they easily look out of place.
  Meaningful item handles are recommended.
  
  ```{r}
  unname(lipset$items["afterlife"])
  ```
  
  If used intext, unnamed items yield a footnote without title such as this unnamed item `r unname(lipset$items["race_4_success"])`.


### Rendering

It is often helpful to have a canonical, typeset version of text items, ready for for printing, web publishing or interpretation.
Rendered text items should meet several criteria:

- For practical reasons, items should fit on the *same* card size.
  To emphasize the equal "significance" of items, their fontsize and design should also be equal.
- They should always *look identical*, no matter where a participant or researcher encounters them.
  Even slight variations in, for example, line wrapping, might give an item a slightly different emphasis.
- Given the central status of text items in the methodology and the package, they should by *typeset professionally*.
 
`psItemContent()` allows you to set the way items are typeset (or rendered) using a variety of options.
This defaults to the settings for printing items on standard-issue business cards, but you can provide arbitrary dimensions and margins, as well as other options.
Just keep in mind that not all options may meet the requirement that your item text *can* in fact be typeset to a *single* page (card), and downstream functions may error out.

You will notice that the `fontsize_global` option in `psItemContent()` does not greatly change results; it should be used for tweaking only.
Instead, pensieve **automatically chooses the largest possible fontsize**, which still allows all items to fit on one page.
This optimisation takes place behind the scenes, and will be run whenever necessary.
However, the process can take several minutes, and will need to rerun whenever you change an item or one of the design parameters.
To cut down on waiting time, you may want to finish all your items before ever rendering them, and rely on the (sensible) default design parameters.
You can always inspect your items by printing them to the screen using the normal `print()` method.

All design options are stored as attributes *with* the `psItemContent` object, and should only be set using the `psItemContent()` function.
For downstream methods and functions, such as `plot()` it is important that items always retain all their attributes in the `psItemContent` object.
Divergent from standard R behavior, these attributes are retained on subsetting using the standard `'['` operator, but may be lost when you otherwise operate on `psItemContent` objects.
You can always reconstruct the object by running `psItemContent()` or `as_psItemContent()` anew.
If you frequently use other operations on `psItemContent` objects and would like to retain attributes, you can suggest it as a feature of pensieve by raising an [issue](https://github.com/maxheld83/pensieve/issues) on github.

In addition to setting design options for *all* items using the arguments in `psItemContent()`, you can also markup *individual* items using (basic) [Pandoc's markdown](https://pandoc.org/MANUAL.html#pandocs-markdown), including *italics*, and many other options.
Just make sure that using Pandoc's markdown markup does not break the constraint of rendering items on a *single* page (or card).[^byoi]

[^byoi]:  If you would like to customize the look of your items even further than that, you can always use a typesetting program of your choice and provide items as binary assets for the `psItemContentBin` class.
  For now, this class has very limited support in the package.

Rendering professionally typeset items, pensieve goes through several steps:

1. Items are converted from Pandoc's Markdown (or plain text) to **LaTeX** via [md2tex_mem()], then
2. compiled to **PDF** via [texi2pdf2_mem()], then
3. converted to **SVG** via [pdf2svg_mem()], then
4. imported to **R Graphics** via [svg2grob_mem()].
   Items are now fully available to the R Graphics system and can be used wherever [graphics::plot()] (or, to be precise, [grid::grid.draw()]) works.

At each step of this necessary, but rather long conversion pipeline more (system) dependencies are required and asserted, so you may need to install additional programs on your computer.
Step 3 (and later) is currently unavailable on machines running Microsoft Windows.

Once the items are rendered, you can:

- `plot()` them using R graphics system.
   This makes the items available in their final, canonical form wherever the R plotting system is.
   R graphics devices must offer arbitrary aspect ratios, while items have a fixed aspect ratio equal to that of the cards.
   As a result, items may be surrounded by additional whitespace in the resulting plots.
   To avoid this, set the aspect ratio of your R graphics to equal that of the cards as seen in the below code chunk.
   
   Because it does not make much sense to `plot()` several items at once, `plot()` will always default to the *first* item.
   You can also subset to your desired item.
   
   ```{r, fig.width = attr(lipset$items, which = "paperwidth"), fig.height = attr(lipset$items, which = "paperheight"), fig.cap = "Item 'Moderation'"}
   #plot(lipset$items["moderation"])
   ```
   
- `output_ps()` them to your file system in one of several vector formats.
   Because some of the intermediary formats cannot be easily or fully converted, downstream outputs may be faulty.
   If you can, always use the *earliest* possible output from the above conversion pipeline to maximize fidelity to the original.


## Bibliography

