---
title: 'Q-Cat: A Method and Algorithm for Categorical Operant Subjectivity'
author: "Verena Kasztantowicz, Maximilian Held"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
bibliography: 
  #- "../../held-library/held_library.bib"
  #- "../../vk_library/verenakasztantowicz.bib"
---

## Abstract

Q-Sorts record item positions *continuously*, distending from a neutral mid-point along some dimension. 
However, while adhering to Stephenson's ontology and epistemology, Q methodology can also be extended to cover *categorical* operant subjectivity.

We here suggest a survey method and analytical algorithm to capture such shared viewpoints of categorizing items.
The well-established [Repertory Grid Technique](https://en.wikipedia.org/wiki/Repertory_grid) also shares some of Q's underpinnings and allows spontaneous categories, but does not lend itself (easily) to a transposed, "Q-wise" factor extraction and may not be ipsative between the items. 
Instead, we invite participants to come up with their own, dichotomous categories and ask them to classify the full Q-Set accordingly.
The resultant logical table is summarized into a per-person measure of categorical item similarity, in turn transformed into a person-by-person correlation matrix.
We then extract higher-level principal components (PCA) in a stepwise residual process, and calculate the respective (rotated) component scores, yielding ideal-typical, shared ways of categorizing the Q-Set.
These categorical viewpoints are then subjected to a nested, low-level PCA to allow a substantive abduction of just *what* those categories might be.

Using real-world data gathered for this purpose, we show that results from the suggested procedure can be meaningfully interpreted as categorical operant subjectivity.
We also present software for the R statistics framework to run the analysis and a web-based interface to gather data.

Categorizing things can be considered a fundamental, spontaneous act of the human mind, and Q-Cat is a helpful addition to the Q-methodologists toolkit to study it.

```{r setup}
library(knitr)
library(rmarkdown)
library(pensieve)
install.packages(repos = NULL, type = "source", INSTALL_opts = c('--no-lock'), pkgs = c("../qmethod"))
```

```{r kill-nas}
data(komki)
komki$qsorts <- na.omit(komki$qsorts)
komki$items <- komki$items[komki$items$Handle.english %in% rownames(komki$qsorts), ]
komki$qcat$ass <- sapply(X = komki$qcat$ass, simplify = FALSE, FUN = function(x) {
  x[rownames(x) %in% rownames(komki$qsorts), , drop = FALSE]
})
lapply(X = komki$qcat$ass, FUN = function(x) {is.matrix(x)})
```


## Creating the Co-Occurence Matrix

```{r make-cooc}
cooc <- count_cooccur(ass = komki$qcat$ass)
```

For example, `Willy`'s co-occurence matrix looks like this:

```{r show-cooc}
kable(cooc[,,"Willy"])
```



## Example Citations

@Adams-2001-aa ist eine Quelle aus Max's bibtex Datei.

@artemeva-fox-2010 ist eine Quelle aus Verenas bibtex Datei.

## Bibliography
