---
title: "qrots"
author: "Maximilian Held"
date: "12 February 2016"
output: pdf_document
---

```{r preamble, message=FALSE, warning=FALSE}
library(devtools)  # just to get the local qmethod package
library(knitr)  # needed for output
library(rmarkdown)  # needed for output
library(grid)  # needed, for, well, grids and arrows
library(gridExtra)  # needed for more grids
library(ggplot2)  # for custom plots
install.packages(repos = NULL, type = "source", INSTALL_opts = c('--no-lock'), pkgs = c("../qmethod"))  # notice this is the LOCAL qmethod
library(qmethod)  # loading it
library(psych)  # to use psych's functions directly
library(reshape2)  # data massaging
library(assertthat)  # testing
```

```{r setup-knitr, cache = FALSE, message=FALSE, warning=FALSE}
he
```

```{r data-import, cache=TRUE}
load("~/Github/qrots/raw.RData")
```
## Analysis

```{r analysis, cache = TRUE}
res <- NULL
res$var <- qmethod(dataset = raw, nfactors = 4, rotation = "varimax", forced = TRUE, cor.method = "spearman", reorder = FALSE, allow.confounded = FALSE)
res$quar <- qmethod(dataset = raw, nfactors = 4, rotation = "quartimax", forced = TRUE, cor.method = "spearman", reorder = FALSE, allow.confounded = FALSE)
res$equa <- qmethod(dataset = raw, nfactors = 4, rotation = "equamax", forced = TRUE, cor.method = "spearman", reorder = FALSE, allow.confounded = FALSE)
# TODO add highlights for some cases
```


## Loadings Plots

Here are the loadings plots in this order: **varimax**, **quartimax** and **equamax**.

```{r loaplots, cache = TRUE, fig.height=10, fig.width=10}
plots <- NULL
plots$loa <- lapply(X = res, FUN = "q.loaplot", names = FALSE, points = TRUE, alpha = 1/3, density = TRUE, grid = FALSE, rug = TRUE, quietly = TRUE, red = FALSE)
for (i in 1:length(plots$loa$var)) {
  do.call(what = "grid.arrange", args = c(plots$loa$var[i], plots$loa$quar[i], plots$loa$equa[i], nrow = 1))
}
```


## Confounded Q-Sorts

These are for the various rotations:

- Varimax: **742**
- Quartimax: **431** (as expected, smaller!)
- Equamax: **431** (same, oddly)


## Number of Flags

The number of flagged q-sorts (with de-confounding) is actually quite similar.

```{r flagging summary, cache = TRUE}
flagsums <- lapply(X = names(res), FUN = function(x) colSums(res[[x]][["flagged"]]))
names(flagsums) <- names(res)
flagsums
lapply(X = flagsums, FUN = "sum")
```


## Loadings Histogram

Again in the order (horizontally) **varimax**, **quartimax** and **equamax**.

We can see that the later factors under quartimax and equamax have quite little variance and mostly low loadings.

```{r loa-histo, cache = TRUE}
library(ggplot2)
for (i in colnames(res$var$loa)) {
  var <- ggplot(data = res$var$loa, mapping = aes_q(x = as.name(i))) + geom_histogram(binwidth = 0.05) + xlim(-1,1)
  quarti <- ggplot(data = res$quar$loa, mapping = aes_q(x = as.name(i))) + geom_histogram(binwidth = 0.05) + xlim(-1,1)
  equa <- ggplot(data = res$equa$loa, mapping = aes_q(x = as.name(i))) + geom_histogram(binwidth = 0.05) + xlim(-1,1)
  grid.arrange(var, quarti, equa, nrow = 1)
}
```

## Summed of Squared Loadings

The variance is redistributed to the first, general factor even *before* flagging:

```{r sum-squares, cache = TRUE}
apply(X = res$var$loa, MARGIN = 2, FUN = function(x) sum(x^2))
apply(X = res$quar$loa, MARGIN = 2, FUN = function(x) sum(x^2))
apply(X = res$equa$loa, MARGIN = 2, FUN = function(x) sum(x^2))
```


